# Introduction

This tutorial shows how to perform power and sample size calculations
in R for continuous outcome variables. For those who are interested, the website below includes examples
for binary outcomes.

Retrieved from: https://publicifsv.sund.ku.dk/~tag/Teaching/share/R-tutorials/Basic-statistics/Power.html

# Continuous outcome

A continuous outcome variable is a variable that can take on any value
between its minimum and its maximum value, such as weight or BMI or
the blood hemoglobin concentration. For the power and sample size
calculations illustrated here the distribution of the outcome variable
is characterized by its mean and standard deviation.

## Power for given sample size

How can we compute the power of a study which aims to show a difference
in means between group 1 (n=6) and group 2 (n=6) assuming that the
magnitude of the difference is 0.3 units and the standard deviation is
0.28 units?

```{r   }
power.t.test(n=6,delta=0.3,sd=0.28,type="two.sample") 
```

Possible conclusion sentence: 

*The power of the study is 39% to detect a difference in means of 0.3 units.*

Is this power good enough? How can we increase the power?

## Sample size for a given power

How can we compute the sample size needed to achieve a power of 90% in a
study which aims to show a difference in means between two independent
groups assuming that the magnitude of the difference is 0.3 units and
the standard deviation is 0.28 units.

```{r   }
# TODO: Use the power.t.test() function to calculate the sample size needed to achieve a power of 90%, assuming that the magnitude of the difference is 0.3 units and the standard deviation is 0.28 units.
help("power.t.test")
```

Possible conclusion sentence: 

*To reach a power of 90% the study should include at least 20 subjects in each group to detect a difference in means of 0.3 units.*

# Plotting power against sample size

The functions `power.prop.test` and `power.t.test` can compute the
power for a sequence of sample sizes simultaneously.

How would we plot the relationship between power and a given sequence of sample sizes for the above example?

```{r   }
samplesizes <- seq(from=10,to=200,by=10)
# TODO: Use the power.t.test() function to compute the power of a sequence of sample sizes. Assume that the magnitude of difference is 0.3 unites and the standard deviation is 0.28 units. Then, use the plot() function to plot that relationship.
```

Often the greatest concern is the magnitude of the expected difference
between the groups, even if based on historical data or a pilot
study. Therefore a useful plot shows how the sample size for fixed
power (or power for fixed sample size) varies as a function of the
difference. However, the functions `power.prop.test` and
`power.t.test` can unfortunately not deal with a sequence of
differences. Therefore we use a loop (`sapply`) in the following
example. To show the effect of the hypothesized standard deviation we
also show the required sample size when the standard deviation is
only 0.2 instead of 0.4.

```{r   }
differences <- seq(from=0.1,to=2,by=0.1)
samplesize_sd04 <- sapply(differences,
                          function(d){power.t.test(power=0.9,
                                                   delta=d,
                                                   sd=0.4,
                                                   type="two.sample")$n})
samplesize_sd03 <- sapply(differences,
                          function(d){power.t.test(power=0.9,
                                                   delta=d,
                                                   sd=0.3,
                                                   type="two.sample")$n})
plot(differences,
     samplesize_sd04,
     xlim=c(0,2),
     xlab="Expected difference between groups",
     ylab="Required sample size",
     ylim=c(0,350),
     type="b",
     col="darkblue",
     lwd=5,axes=FALSE)
lines(differences, samplesize_sd03,col="turquoise",lwd=5,type="b")
axis(1,at=c(0,0.2,0.5,1,1.5,2))
axis(2,at=c(350,100,50,10,0))
legend(x="topright",lwd=5,bty="n",legend=c("SD=0.4","SD=0.3"),col=c("darkblue","turquoise")) 
```
