---
title: "2022_ANOVA_script"
output: html_document
date: '2022-06-20'
---

# Load up dplyr
```{r}
# We are going to need to load a package called dplyr for this. If you do not have 
# the dplyr package, then please run install.packages("dplyr") in the console first.
library(dplyr)
```

# Generate some data
Let's generate some data using three different calls to the rnorm function to represent samples from three different populations.
```{r}
# Use a sample size of 20 for all of the samples. Set the mean of the first two samples as 2 and set the 
# mean of the last sample as 4. Use a standard deviation of 1.5 for all samples. We are going to call
# the three samples x1, x2, and x3. 
x1 = rnorm(20, mean = 2, sd = 1.5)
x2 = rnorm(20, mean = 2, sd = 1.5)
x3 = rnorm(20, mean = 4, sd = 1.5)
# Let's look at the mean and spread of each of our samples using the boxplot() function. 
boxplot(x1,x2,x3)
```

# Organize Your Data
The ANOVA function is a little bit different from the other functions in that it works best when your data is organized a certain way. Luckily, this is the way your data will usually be organized after collection anyway.
```{r}
# Let's bind the example data together using cbind() and then put it into a dataframe with data.frame(). 
# Let's call our new dataframe data. 
data = data.frame(cbind(x1,x2,x3))

# Take a look at the dataframe we just made. Right now we have each group organized into a different 
# column. This format is easy for a person to understand but hard for a computer to understand. It is not
# tidy. We need to organize it so that all of the sample values are in one column and the group each 
# value belongs to is in a 2nd column. We can do this very easily with the stack() function which 
# automatically organizes our data this way. 
data = stack(data)
  
# look at our reorganized dataframe and think about how it has changed. 
View(data)  
  
# The last thing we are going to do is use dplyr's rename() function to change the column name "ind" to 
# "group".
data = rename(data, group = ind)

# Now our data is all set for ANOVA!
```

# ANOVA
we are going to learn ANOVA in a kind of backwards way. We are going to start by actually conducting an ANOVA test and then we are going to use R to explore what all of the different outputs mean. 

```{r}
# In order to run an ANOVA test, we are going to need to use R's linear regression function lm()
# Why are we using linear regression for ANOVA? Because ANOVA is actually a special application of linear
# regression! We may explore this more in future workshops, but for now do not worry about it. 

# Run help(lm). Focus on the first two arguments. 
help(lm)

# Wait, what is a formula class object? 
# Formula class objects are specified using a tilde (~) and are used to specify a relationship between 
# two vectors.
# Essentially, when you type A~B you are saying that you want to test if variable A is dependent on B.
# The first term is the response variable in your data and the second is the explanatory variable. 
# The specifics of what a formula object actually is in R are complicated and beyond this workshop. 
# If you want more details, look up ?formula after the workshop.  

# Ok, now we can actually use lm() by specifying the formula and data arguments. Run lm and save the 
# output in a variable named Av_object
# Hint: what are our response and explanatory variables? Specify the variables using the column names.
Av_object = lm(values~group,data = data) 

# Now we are ready to actually run the ANOVA test using the anova() function. 
anova(Av_object)

```
